Storm surge modelling in the Strait of Georgia

### 2014-11-05 RoughWork/manuscript.tex
Fixed up models paragraph. Still need to add Giddings ref.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. \citet{yang2010multi} developed a model of southern Strait of Georgia and Puget Sound using finite volume techniques on an unstructured grid. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  These studies have led to an improvement in understanding circulation properties in the Salish Sea and have highlighted important features that should be considered in future modelling work. 
%What was interesting about Yang et al?
%Tides in Sutherland? JdF surface salinity in Giddings
%Add more about reproduction of tides in other models 
%Last sentance is lame

@@

%storm surge modelling
%murty, danard, crean, non BC models (xu, flather)
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{Model domain (highlighted area) including bathymetry, rivers (blue circles), and storm surge locations of interest  (red stars).}\label{fig:domain}
\end{figure}
%image may need more work
%Add Johnstone Strait, Vancouver, Fraser River, Haro Strait, San Juan/Gulf Islands

@@

\subsection{Example Storm Surge Hindcast: Feb 4, 2006}

@@

@@

\subsection{Spatial Extent of Surge: Dec 15, 2006}
On Dec 15, 2006, a major storm with wind gusts over 120 km/hr caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. This event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\subsection{Strong Wind Event: Nov 19, 2009}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Historical Climate Data archives, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases as shown in Figure \ref{fig:nov2009} where we have plotted the wind speed at Point Atkinson, sea surface height anomaly forcing at the open boundaries, and observed and modelled residuals at Point Atkinson. Two significant surge events are noted, the first occurring on Nov 19 and a second, larger surge on Nov 20. The largest surge did not coincide with the strong wind event which peaked at Nov 19 01:00 UTC, reflected in both the model and observations. At Point Atkinson, the highest observed surge reached 62.5 cm on Nov 20 at 10:30 UTC, while the model recorded a maximum surge of 50.5 cm five hours later. On Nov 19, the observed surge reached upwards of 50 cm, which is low when compared to the other cases considered. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy. 
%citation for Environment Canada data?

@@

\subsection{An Additional Hindcast: Nov 16, 2006}
An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast, Willmott skill scores less greater than 0.94 as outlined in Table \ref{tab:nov2006stat}.  The best match in maximum surge occured at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and slightly earlier at the other stations.

>>>>

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months \citep{thomson1981oceanography}. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation \citep{masson2004modelling}. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. A finite volume model on an unstrcutured grid by \citet{yang2010multi} used a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semidiurnal amplitudes were generally too low. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases is a recurring difficulty in modelling of this region.
% JdF surface salinity in Giddings
%Add more about reproduction of tides in other models

@@

%storm surge modelling
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{Model domain (highlighted area) including bathymetry, rivers (blue circles), and storm surge locations of interest  (red stars).}\label{fig:domain}
\end{figure}

@@

\subsection{Example Storm Surge Hindcasts}

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast, Willmott skill scores are greater than 0.94 as outlined in Table \ref{tab:nov2006stat}.  The best match in maximum surge occured at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and slightly earlier at the other stations.

@@

\subsection{Spatial Extent of Surge}
On Dec 15, 2006, a major storm with wind gusts over 120 km/hr caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. This event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\subsection{Strong Wind Event}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Historical Climate Data archives, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases as shown in Figure \ref{fig:nov2009} where we have plotted the wind speed at Point Atkinson, sea surface height anomaly forcing at the open boundaries, and observed and modelled residuals at Point Atkinson. Two significant surge events are noted, the first occurring on Nov 19 and a second, larger surge on Nov 20. The largest surge did not coincide with the strong wind event which peaked at Nov 19 01:00 UTC, reflected in both the model and observations. At Point Atkinson, the highest observed surge reached 62.5 cm on Nov 20 at 10:30 UTC, while the model recorded a maximum surge of 50.5 cm five hours later. On Nov 19, the observed surge reached upwards of 50 cm, which is low when compared to the other cases considered. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy. 
%citation for Environment Canada data?

@@
